DRAFT: Direct-style process creation on Linux

### 2020-05-16 research/paper/paper.tex
paper: remove mention of thread

We use the term ``thread'' to refer to the processes under the control of \texttt{rsyscall},
including the ``local'' process.
On Linux, the shared-memory ``threads'' provided by libraries such as \texttt{pthreads}
are implemented as processes,
and their lifetime and execution is completely controlled by a single program.
The same is mostly true of the controlled-process ``threads'' provided by \texttt{rsyscall}.
The most significant difference is that \texttt{rsyscall} ``threads''
do not run user code concurrently with the main program;
nevertheless \texttt{rsyscall} ``threads'' do provide the opportunity for concurrent execution of system calls,
and so the terminology provides useful intuition.

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' thread to create a new child thread.
We call various syscalls in the child thread to mutate it until it reaches the desired state,
at which point we can call \texttt{execve} in the child thread to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process
(hence we say ``process'' rather than ``thread'').
The child process can be monitored using normal Linux child monitoring syscalls from our main thread,
such as \texttt{waitid},
which we can also do with \texttt{rsyscall}.

\lstinputlisting[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
{examples/basic/direct.py}
Listing \ref{basic} shows this in action in Python.
We create a new child thread using \texttt{clone} on \verb|local_thread|,
and receive back a capability to control it.
We pass no flags to use \texttt{clone}'s defaults
(which essentially match \texttt{fork}) to create the process.
We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child thread capability will fail with an exception.

@@

\lstinputlisting[float=*,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child thread,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
\verb|inherit_fd| only updates resource-tracking bookkeeping in our program,
it makes no system calls;
the file descriptor was already copied at process creation time.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
so that we don't need to pass the (unmodified) environment to every exec call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

@@

\lstinputlisting[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
{examples/mount/direct.py}
One classic form of customization enabled by namespaces
is overriding files at paths that were hardcoded into a program we can't change,
without changing those files for the rest of the system.
In listing \ref{mount},
we create a new child thread as normal,
but for the first time,
pass an argument to \texttt{clone}.

@@

\lstinputlisting[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
{examples/unmount/direct.py}
In listing \ref{unmount}
we create a new child thread,
again in a new mount namespace using \texttt{CLONE.NEWNS}.
We open the executable that we'll later exec in this child thread,
as well as another file as in listing \ref{fds},
since we won't be able to do exec or open files normally after the next step.
Then we use \texttt{umount},
passing \texttt{MNT.DETACH} to perform a recursive unmount of the entire filesystem tree,
removing it all from the view of this process.
As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
so that it can be inherited across \texttt{exec};
we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.

@@

To create another process in the pid namespace,
we clone again, this time from \texttt{init}.
This is the first example in which we've cloned from one of our child threads;
as normal, this produces a new process, a grandchild,
which can be used exactly like a direct child thread.
We exec in the grandchild,
and we can monitor the grandchild process from \texttt{init},
just as we would monitor \texttt{init} from its parent process.

@@

The most simple way to access memory is for the main thread to be in the same address space as the target process.
This is the case most of the time; we pass \texttt{CLONE.VM} by default in the Python wrapper.
However, the target process may be in another address.
For example, if the target process is at a different privilege level,
it must be in a different address space,
as sharing an address space with a process at a different privilege level is unsafe.

>>>>

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
We call various syscalls in the child process to mutate it until it reaches the desired state,
at which point we can call \texttt{execve} in the child process to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process.
The child process can be monitored using normal Linux child monitoring syscalls,
such as \texttt{waitid}.

\lstinputlisting[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
{examples/basic/direct.py}
Listing \ref{basic} shows this in action in Python.
We create a new child process using \texttt{clone} on the local process \verb|local|,
and receive back a capability to control it.
We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
Then we call \texttt{execve},
passing the path of the \texttt{cat} executable,
the arguments to the new executable,
and the unchanged set of environment variables.
The name of the executable is passed as the first argument,
as is typical in Unix.
Calling \texttt{execve} consumes the capability,
and later use of this child process capability will fail with an exception.

@@

\lstinputlisting[float=*,language=Python,label={fds},caption={Passing down FDs}]
{examples/fds/direct.py}
In Listing \ref{fds},
we first open a file with read-write permission.
Then we create a child process,
which inherits all file descriptors from its parent.
We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
\verb|inherit_fd| only updates resource-tracking bookkeeping in our program,
it makes no system calls;
the file descriptor was already copied at process creation time.
Then we disable \texttt{CLOEXEC} on the child's file descriptor,
which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
Finally, we execute some executable,
passing the file descriptor number as an argument;
when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
so that we don't need to pass the (unmodified) environment to every exec call in these examples.
\subsection{Non-shared-memory concurrency}
Processes run concurrently,
which allows exploiting the parallelism of the underlying hardware.
Since processes don't share memory,
they can provide a less complex parallel programming environment
than shared-memory thread-based approaches.

@@

\lstinputlisting[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
{examples/mount/direct.py}
One classic form of customization enabled by namespaces
is overriding files at paths that were hardcoded into a program we can't change,
without changing those files for the rest of the system.
In listing \ref{mount},
we create a new child process as normal,
but for the first time,
pass an argument to \texttt{clone}.

@@

\lstinputlisting[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
{examples/unmount/direct.py}
In listing \ref{unmount}
we create a new child process,
again in a new mount namespace using \texttt{CLONE.NEWNS}.
We open the executable that we'll later exec in this child process,
as well as another file as in listing \ref{fds},
since we won't be able to do exec or open files normally after the next step.
Then we use \texttt{umount},
passing \texttt{MNT.DETACH} to perform a recursive unmount of the entire filesystem tree,
removing it all from the view of this process.
As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
so that it can be inherited across \texttt{exec};
we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.

@@

To create another process in the pid namespace,
we clone again, this time from \texttt{init}.
This is the first example in which we've cloned from one of our child processes;
as normal, this gives us a capability for a new process, a grandchild,
which can be used exactly like a direct child process.
We exec in the grandchild,
and we can monitor the grandchild process from \texttt{init},
just as we would monitor \texttt{init} from its parent process.

@@

The most simple way to access memory is for the local process to be in the same address space as the target process.
This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
However, the target process may be in another address.
For example, if the target process is at a different privilege level,
it must be in a different address space,
as sharing an address space with a process at a different privilege level is unsafe.
